Studies of quantum dots

### 2016-11-02 Manuscript/paper.tex
Remove incorrect factor of 4 in spEnergiesQD/Manuscript/paper.tex
Update spEnergiesQD/Manuscript/paper.tex

We model the quantum dot as a collection of $N$ electrons trapped in an
external harmonic-oscillator potential.\footnote{In this paper, we use atomic
  units where $\hbar$, $m$, $e$, and $4 \pi \epsilon$ are set to 1.  Here,
  $\epsilon$ denotes the permittivity and $m$ denotes the mass, which could be
  an effective mass rather than the true electron mass.  In effect, the unit
  of energy in this article is
  $E_{\text h} (m / m_{\text e}) / (\epsilon / \epsilon_0)^2$ where
  $E_{\text h} \equiv m_{\text e} (\alpha c)^2$, the hartree unit of energy.}
Each electron possesses kinetic energy $t(\bm p) \equiv \bm p^2 / 2$ and
experiences an external potential $u(\bm r) \equiv \omega^2 \bm r^2 / 2$,
where $\omega$ is the angular frequency of the oscillator, $\bm r$ is the
2-dimensional position vector of an electron, and $\bm p$ is its linear
momentum.  Each pair of electrons experience a Coulomb repulsion
$v(\bm r, \bm r') \equiv |\bm r - \bm r'|^{-1}$ where $\bm r$ and $\bm r'$ are
their positions.  The full many-body problem is described by the
nonrelativistic Hamiltonian
\begin{align*}
  \hat H \equiv \hat H^{\text o} + \hat V = \sum_{p} \bigl(
    t(\hat{\bm p}_p) + u(\hat{\bm r}_p)
  \bigr) + \sum_{p < q} v(\hat{\bm r}_p, \hat{\bm r}_q)
\end{align*}
where $\hat{\bm p}_p$ and $\hat{\bm r}_p$ respectively denote the momentum and
position operators of a particle $p$.  The first summation is performed over
every particle, while the the second summation is performed over every
unordered pair of distinct particles (the ordering of $p$ is arbitrary).

The full problem is not a trivial one.  We shall first consider a much simpler problem: the noninteracting part of the Hamiltonian, $\hat H^{\text o}$, which does not contain the Coulomb interaction $v$:
\begin{align} \label{eq:onebodyhamiltonian}
  \hat H^{\text o} \equiv \sum_p \hat h^{\text o}_p
  \equiv \sum_{p} \bigl(
  t(\hat{\bm p}_p) + u(\hat{\bm r}_p)
  \bigr)
\end{align}
In this case, the many-body Hamiltonian reduces to $N$ independent
single-particle Hamiltonians, each of the form:
\begin{align*}
  \hat h^{\text o} \equiv
  \frac{\hat{\bm p}^2 + \omega^2 \hat{\bm x}^2}{2}
\end{align*}
This is a 2-dimensional harmonic oscillator problem for a single particle,
whose solutions are well known.  There are various ways to construct the set
of eigenfunctions: we shall use the Fock-Darwin states that conserve angular
momentum $\hat{L}_{\text{z}}$ about the $z$-axis (perpendicular to the 2D
plane).  The eigenfunctions have the following analytic form in polar
coordinates:\citep[Appx.\ A]{lohne2010coupled}
\begin{gather*}
  \phi^{\text{FD}}_{n, m_\ell}(r, \varphi)
  \equiv \sqrt\omega
  \phi^{\text{FD.R}}_{n, |m_\ell|}(\sqrt \omega r)
  \phi^{\text{FD.A}}_{m_\ell}(\varphi)
\end{gather*}
where
\begin{align*}
  \phi^{\text{FD.R}}_{n, \mu}(\rho) &\equiv
  \sqrt{\frac{2 \cdot n!}{(n + \mu)!}}
  {\mathrm e}^{- \rho^2 / 2} \rho^\mu L_n^{(\mu)}(\rho^2) \\
  \phi^{\text{FD.A}}_{m_\ell}(\varphi) &\equiv
  \frac{1}{\sqrt{2 \pi}} {\mathrm e}^{i m_\ell \varphi}
\end{align*}
and $L_n^{(\mu)}$ denotes the generalized Laguerre polynomial of degree $n$
with parameter $\mu$.  The states are labeled by two quantum numbers: the
principal quantum number $n \in \{0, 1, \ldots\}$ and the angular momentum
projection $m_\ell \in \mathbb Z$.

Additionally, since electrons are spin-$1/2$ fermions, they can occupy either
of the two spin states $\chi_{-\frac12}$ or $\chi_{+\frac12}$.  Thus, the
single-particle basis states can be constructed as
\begin{align} \label{eq:singleparticlestate}
  \phi_{n, m_\ell, m_{\text s}}(r, \varphi) \equiv
  \phi^{\text{FD}}_{n, m_\ell}(r, \varphi) \chi_{m_{\text s}}
\end{align}
with the spin projection $m_{\text s} \in \bigl\{-\frac12, +\frac12\bigr\}$.

@@

The matrix elements of the generator can be evaluated using \eqref{eq:commutmxe} to \eqref{eq:flow2}.  If define $\hat{H}^{\text{d}}$ via
\begin{align*}
  H^{\text{d}, [1]}_{p, q} &\equiv \delta_{p, q} H^{[1]}_{p, p} \\
  H^{\text{d}, [2]}_{p, q, r, s} &\equiv (\delta_{p, r} \delta_{q, s} - \delta_{p, s} \delta_{q, r}) H^{[2]}_{p, q, p, q}
\end{align*}
then the expressions for $\hat{\eta}^{\text{Wg}}$ can be simplified considerably:
\begin{align*}
  \eta^{\text{Wg},[0]} &= 0 \\
  \eta^{\text{Wg},[1]}_{p, q} &= 2 H^{[1]}_{p, q} \mathcal{A}_{p, q} \bigl( H^{[1]}_{p, p} + n_p H^{[2]}_{p, q, p, q} \bigr) \\
  \eta^{\text{Wg},[2]}_{p, q, r, s} &= 2 \mathcal{A}_{p, q} \mathcal{A}_{r, s} \Big( \\
  &\qquad H^{[2]}_{p, q, r, s} \mathcal{A}_{(p, q), (r, s)} \Bigl(\\
  &\qquad \qquad 2 H^{[1]}_{p, p} + 4 n_p H^{[2]}_{p, r, p, r} \\
  &\qquad \qquad + (1 - 2 n_p) H^{[2]}_{p, q, p, q} \Bigr) \\
  &\qquad + 4 \delta_{p, r} H^{[1]}_{q, s} \mathcal{A}_{(p, q), (r, s)} H^{[2]}_{p, q, p, q}
  \Bigr)
\end{align*}
where for any $x$, $y$, $z$, $w$, $f$ we define the antisymmetrizer
\begin{align*}
  \mathcal{A}_{(x, y), (z, w)} f(x, y, z, w) &\equiv \frac{1}{2} \bigl(f(x, y, z, w) - f(z, w, x, y)\bigr)
\end{align}

@@

The generator is explicitly constructed the following way \cite{PhysRevLett.106.222502,White:cond-mat0201346}
\begin{align}
\hat{\eta}^{\text{Wh}} &\equiv \hat{\eta}' - \hat{\eta}'{}^\dagger
\end{align}
where
\begin{align}
\eta^{\prime [1]}_{a, i} &\equiv \frac{\bar{n}_a n_i H^{[1]}_{a, i}}{H^{[1]}_{a, a} - H^{[1]}_{i, i} - H^{[2]}_{a, i, a, i}}
\label{eq:WhiteFull1} \\
\eta^{\prime [2]}_{a, b, i, j} &\equiv \frac{4 \bar{n}_a \bar{n}_b n_i n_j H^{[2]}_{a, b, i, j}}{H^{[1]}_{a, a} + H^{[1]}_{b, b} - H^{[1]}_{i, i} - H^{[1]}_{j, j} + X_{a, b, i, j}}
\label{eq:WhiteFull2} \\
X_{a, b, i, j}
  &\equiv H^{[2]}_{a, b, a, b} - H^{[2]}_{a, i, a, i} - H^{[2]}_{b, i, b, i} \notag \\
  &\qquad + H^{[2]}_{i, j, i, j} - H^{[2]}_{a, j, a, j} - H^{[2]}_{b, j, b, j}
\label{eq:White7}
\end{align}

>>>>

We model the quantum dot as a collection of $N$ electrons trapped in an
external harmonic-oscillator potential.\footnote{In this paper, we use atomic
  units where $\hbar$, $m$, $e$, and $4 \pi \epsilon$ are set to 1.  Here,
  $\epsilon$ denotes the permittivity and $m$ denotes the mass, which could be
  an effective mass rather than the true electron mass.  In effect, the unit
  of energy in this article is
  $E_{\text h} (m / m_{\text e}) / (\epsilon / \epsilon_0)^2$ where
  $E_{\text h} \equiv m_{\text e} (\alpha c)^2$, the hartree unit of energy.}
Each electron possesses kinetic energy $t(\bm p) \equiv \bm p^2 / 2$ and
experiences an external potential $u(\bm r) \equiv \omega^2 \bm r^2 / 2$,
where $\omega$ is the angular frequency of the oscillator, $\bm r$ is the
2-dimensional position vector of an electron, and $\bm p$ is its linear
momentum.  Each pair of electrons experience a Coulomb repulsion
$v(\bm r, \bm r') \equiv |\bm r - \bm r'|^{-1}$ where $\bm r$ and $\bm r'$ are
their positions.  The full many-body problem is described by the
nonrelativistic Hamiltonian
\begin{align*}
  \hat H \equiv \hat H^{\text o} + \hat V = \sum_j \bigl(
    t(\hat{\bm p}_j) + u(\hat{\bm r}_j)
  \bigr) + \sum_{j_1 < j_2} v(\hat{\bm r}_{j_1}, \hat{\bm r}_{j_2})
\end{align*}
where $\hat{\bm p}_j$ and $\hat{\bm r}_j$ respectively denote the momentum and
position operators of a particle $j$.  The first summation is performed over
every particle, while the the second summation is performed over every
unordered pair of distinct particles (the ordering of the particle labels $j$ is arbitrary).

The full problem is not a trivial one.  We shall first consider a much simpler problem: the noninteracting part of the Hamiltonian, $\hat H^{\text o}$, which does not contain the Coulomb interaction $v$:
\begin{align} \label{eq:onebodyhamiltonian}
  \hat H^{\text o} \equiv \sum_p \hat h^{\text o}_p
  \equiv \sum_{p} \bigl(
  t(\hat{\bm p}_p) + u(\hat{\bm r}_p)
  \bigr)
\end{align}
In this case, the many-body Hamiltonian reduces to $N$ independent
single-particle Hamiltonians, each of the form:
\begin{align*}
  \hat h^{\text o} \equiv
  \frac{\hat{\bm p}^2 + \omega^2 \hat{\bm r}^2}{2}
\end{align*}
This is a 2-dimensional harmonic oscillator problem for a single particle,
whose solutions are well known.  There are various ways to construct the set
of eigenfunctions: we shall use the Fock-Darwin states that conserve angular
momentum $\hat{L}_{\text{z}}$ about the $z$-axis (perpendicular to the 2D
plane).  The eigenfunctions have the following analytic form in polar
coordinates:\citep[Appx.\ A]{lohne2010coupled}
\begin{gather*}
  \phi^{\text{FD}}_{n, m_\ell}(\rho, \varphi)
  \equiv \sqrt\omega
  \phi^{\text{FDr}}_{n, |m_\ell|}(\sqrt \omega \rho)
  \phi^{\text{Lz}}_{m_\ell}(\varphi)
\end{gather*}
where $\rho$ denotes the radial coordinate, $\varphi$ denotes the angular coordinate,
\begin{align*}
  \phi^{\text{FDr}}_{n, m}(r) &\equiv
  \sqrt{\frac{2 \cdot n!}{(n + m)!}}
  {\mathrm e}^{-r^2 / 2} r^m L_n^{(m)}(r^2) \\
  \phi^{\text{Lz}}_{m_\ell}(\varphi) &\equiv
  \frac{1}{\sqrt{2 \pi}} {\mathrm e}^{i m_\ell \varphi}
\end{align*}
and $L_n^{(\mu)}$ denotes the generalized Laguerre polynomial of degree $n$
with parameter $\mu$.  The states are labeled by two quantum numbers: the
principal quantum number $n \in \{0, 1, \ldots\}$ and the angular momentum
projection $m_\ell \in \mathbb Z$.

Additionally, since electrons are spin-$1/2$ fermions, they can occupy either
of the two spin states $\chi_{-\frac12}$ or $\chi_{+\frac12}$.  Thus, the
single-particle basis states can be constructed as
\begin{align} \label{eq:singleparticlestate}
  \phi_{n, m_\ell, m_{\text s}}(\rho, \varphi) \equiv
  \phi^{\text{FD}}_{n, m_\ell}(\rho, \varphi) \chi_{m_{\text s}}
\end{align}
with the spin projection $m_{\text s} \in \bigl\{-\frac12, +\frac12\bigr\}$.

@@

The matrix elements of the generator can be evaluated using \eqref{eq:commutmxe} to \eqref{eq:flow2}.  If define $\hat{H}^{\text{d}}$ via
\begin{align*}
  H^{\text{d}, [1]}_{p, q} &\equiv \delta_{p, q} H^{[1]}_{p, p} \\
  H^{\text{d}, [2]}_{p, q, r, s} &\equiv (\delta_{p, r} \delta_{q, s} - \delta_{p, s} \delta_{q, r}) H^{[2]}_{p, q, p, q}
\end{align*}
then the expressions for $\hat{\eta}^{\text{Wg}}$ can be simplified considerably:
\begin{align*}
  \eta^{\text{Wg},[0]} &= 0 \\
  \eta^{\text{Wg},[1]}_{p, q} &= 2 H^{[1]}_{p, q} \mathcal{A}_{p, q} \bigl( H^{[1]}_{p, p} + n_p H^{[2]}_{p, q, p, q} \bigr) \\
  \eta^{\text{Wg},[2]}_{p, q, r, s} &= 2 \mathcal{A}_{p, q} \mathcal{A}_{r, s} \Big( \\
  &\qquad H^{[2]}_{p, q, r, s} \mathcal{A}_{(p, q), (r, s)} \Bigl(\\
  &\qquad \qquad 2 H^{[1]}_{p, p} + 4 n_p H^{[2]}_{p, r, p, r} \\
  &\qquad \qquad + (1 - 2 n_p) H^{[2]}_{p, q, p, q} \Bigr) \\
  &\qquad + 4 \delta_{p, r} H^{[1]}_{q, s} \mathcal{A}_{(p, q), (r, s)} H^{[2]}_{p, q, p, q}
  \Bigr)
\end{align*}
where for any $x$, $y$, $z$, $w$, $f$ we define the antisymmetrizer
\begin{align*}
  \mathcal{A}_{(x, y), (z, w)} f(x, y, z, w) &\equiv \frac{1}{2} \bigl(f(x, y, z, w) - f(z, w, x, y)\bigr)
\end{align*}

@@

The generator is explicitly constructed the following way \cite{PhysRevLett.106.222502,White:cond-mat0201346}
\begin{align}
\hat{\eta}^{\text{Wh}} &\equiv \hat{\eta}' - \hat{\eta}'{}^\dagger
\end{align}
where
\begin{align}
\eta^{\prime [1]}_{a, i} &\equiv \frac{\bar{n}_a n_i H^{[1]}_{a, i}}{H^{[1]}_{a, a} - H^{[1]}_{i, i} - H^{[2]}_{a, i, a, i}}
\label{eq:WhiteFull1} \\
\eta^{\prime [2]}_{a, b, i, j} &\equiv \frac{\bar{n}_a \bar{n}_b n_i n_j H^{[2]}_{a, b, i, j}}{H^{[1]}_{a, a} + H^{[1]}_{b, b} - H^{[1]}_{i, i} - H^{[1]}_{j, j} + X_{a, b, i, j}}
\label{eq:WhiteFull2} \\
X_{a, b, i, j}
  &\equiv H^{[2]}_{a, b, a, b} - H^{[2]}_{a, i, a, i} - H^{[2]}_{b, i, b, i} \notag \\
  &\qquad + H^{[2]}_{i, j, i, j} - H^{[2]}_{a, j, a, j} - H^{[2]}_{b, j, b, j}
\label{eq:White7}
\end{align}
